---
title: "Species change - VOCC data"
format: 
  html:
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(here)
library(iucnredlist) ### remotes::install_github("IUCN-UK/iucnredlist")
library(janitor)
library(cowplot)
```

This document investigates the Sorensen Similarity Index for bird species in the Lawler paper at both the species and trait level. However, instead of the Lawler projections, we use updated future species ranges created from the VoCC package. Visualizations are at the pixel level. Categorical trait data is from AVONET. IUCN threat data was created in 9_lawler_iucn_threat_status.qmd

```{r}
# Functions 
source(here("common_fxns.R"))
```

```{r}
# Read in VoCC projected future species range
spp_mapfiles <- list.files(here_anx('/spp_range_proj_cordex/spp_range_projected'), 
                           full.names = TRUE, recursive = TRUE, pattern = '.tif')
spp_all_df <- data.frame(f = spp_mapfiles,
                         spp_rcp = basename(spp_mapfiles) %>%
                           str_remove('_range') %>%
                           str_remove('_2060.+') %>%
                           str_replace_all('_', ' ')) %>% 
  mutate(spp = str_sub(spp_rcp, 1, -5),
         rcp = str_sub(spp_rcp, -3, -1))
```

```{r}
# This script focuses on the RCP 8.5 projections
spp_rcp85 <- spp_all_df %>% 
  filter(rcp == "8.5") %>% 
  dplyr::select(!spp_rcp)
```

```{r}
# Read in historical species range (same Lawler species ranges but updated to CORDEX resolution and crs)
spp_mapfiles_hist <- list.files(here_anx('/spp_range_proj_cordex/spp_clim_hist'), 
                           full.names = TRUE, recursive = TRUE, pattern = '.tif')
spp_all_df_hist <- data.frame(f = spp_mapfiles_hist,
                         spp = basename(spp_mapfiles_hist) %>%
                           str_remove('clim_hist_') %>%
                           str_remove('.tif') %>% 
                           str_replace_all('_', ' ')) 
```

```{r}
# Read in AVONET dataset (reading in BirdLife tab, which should match with IUCN names)
avonet <- read_csv(here("data-laa", "AVONET-birdlife-tab-supplementary-dataset-1.csv")) %>% 
  clean_names()
```

```{r}
# Read in the IUCN threat status data
iucn_threat <- read_csv(here("data-laa", "lawler_birds_iucn_status.csv"))
```

## Gather maps and aggregate

Following the method in 2_combine_spp_scenarios.qmd: 
For all the Lawler birds species (301) gather the future maps (just one per species unlike the Lawler data). Then crop the historical maps to the same extent. 

### Aggregate future projections

For each species in the sample, pull the future scenarios.  Crop to the lower continental US states.

### crop future maps

```{r}
### set up extent for continental US
us_extent <- ext(c(xmin = -130, xmax = -65, ymin = 23, ymax = 50))

### Aggregate over the vector of species names
spp_vec <- spp_rcp85$spp %>% unique()
```

```{r}
future_rast_vocc <- lapply(spp_vec, FUN = agg_scenario, df = spp_rcp85) %>%
  setNames(spp_vec) %>%
  rast()

# Test out one of the maps
x <- future_rast_vocc[[16]]
plot(x, main = names(future_rast_vocc)[16])

```

### crop historical maps (these are the Lawler historical ranges, but updated to CORDEX data resolutionand crs)

```{r}
spp_vec_hist <- spp_all_df_hist$spp %>% unique()
```

```{r}
hist_rast <- lapply(spp_vec_hist, FUN = agg_scenario, df = spp_all_df_hist) %>%
  setNames(spp_vec_hist) %>%
  rast()

x_hist <- hist_rast[[16]]
plot(x_hist, main = names(hist_rast)[16])

```

## Generate jurisdiction map

Let's generate a simple jurisdiction map - US states, and Canada and Mexico.

```{r}
states_sf <- ne_states(returnclass = 'sf', country = c('Mexico', 'Canada', 'United States of America')) 

state_id_df <- states_sf %>%
  st_drop_geometry() %>%
  mutate(juris = ifelse(adm0_a3 == 'USA', name, admin)) %>%
  dplyr::select(id = gn_id, juris)
### plot(states_sf %>% select(iso_a2))
states_rast <- rasterize(states_sf, hist_rast, field = 'gn_id')
#plot(states_rast)
```

```{r}
# Function to get pixels instead of jurisdictions
extract_jurisdictions_pixels <- function(spp, future_r, hist_r, juris_r) {
  ### spp <- spp_vec[1]
  ### future_r = future_rast; hist_r = hist_rast; juris_r = states_rast
  s_f_r <- future_r[[names(future_r) == spp]]
  s_h_r <- hist_r[[names(hist_r) == spp]]
  s_df <- data.frame(values(s_h_r),
                     values(s_f_r),
                     values(juris_r)) %>%
    setNames(c('historical', 'future', 'id')) %>%
    mutate(pixel_id = row_number()) %>% 
    filter(!is.na(id)) %>%
    filter(!is.na(historical) | !is.na(future)) #%>%
    #group_by(id) %>%
    #summarize(n_historical = sum(historical > .25),
              #n_future = sum(future > .5))

  return(s_df)
}
```

## Rough pass at classification?

Assign any pixel greater than zero to presence (1). Count up the spp ineach pixel

```{r}
pixel_spp_list <- lapply(spp_vec, FUN = extract_jurisdictions_pixels,
                         future_r = future_rast_vocc, 
                         hist_r = hist_rast,
                         juris_r = states_rast) %>%
  setNames(spp_vec) %>%
  bind_rows(.id = 'spp') %>%
  left_join(state_id_df, by = 'id') #%>%
  ### note, Canada and Mexico have multiple jurisdictions (provinces and states)
  ### so sum up all those instances
  #group_by(spp, juris) %>%
  #summarize(n_historical = sum(n_historical),
            #n_future = sum(n_future), 
            #.groups = 'drop')
```

```{r}
# Assign 1 to values greater than zero
pixel_spp_list_presence <- pixel_spp_list %>% 
  mutate(historical = ifelse(historical > 0, 1, historical),
         future = ifelse(future > 0, 1, future))

# Count up species by pixel
pixel_spp_count <- pixel_spp_list_presence %>% 
  group_by(pixel_id, juris) %>% 
  summarize(historic_spp_count = sum(historical, na.rm = TRUE),
            future_spp_count = sum(future, na.rm = TRUE))
```

## Pixel map

```{r}
# Create a blank raster with same extent as x = spp_rcp85[[16]] 
id_rast <- rast(x)
```

```{r}
# Assign IDs and then convert to sf
values(id_rast ) <- 1:ncell(x)
```

```{r}
id_rast_sf <- id_rast %>% 
  stars::st_as_stars() %>% 
  st_as_sf() %>% 
  rename("pixel_id" = "anas fulvigula")
```

```{r}
# Merge in species count
pixel_spp_count_sf <- inner_join(id_rast_sf, pixel_spp_count)
```

```{r}
historic_spp_num <- ggplot() +
  geom_sf(data = pixel_spp_count_sf, aes(geometry = geometry, fill = historic_spp_count, color = historic_spp_count)) + 
  scale_fill_viridis_c(option = "plasma", limits = c(1,180)) + 
  scale_color_viridis_c(option = "plasma", limits = c(1,180)) +
  geom_sf(data = pixel_spp_count_sf %>% filter(historic_spp_count == 0), aes(geometry = geometry), fill = "gray40", color = "grey40") + 
  labs(title = "VoCC historic number of bird species",
       subtitle = "(0 species in gray)",
       fill = "count",
       color = "count")
```

```{r}
future_spp_num <- ggplot() +
  geom_sf(data = pixel_spp_count_sf, aes(geometry = geometry, fill = future_spp_count, color = future_spp_count)) + 
  scale_fill_viridis_c(option = "plasma", limits = c(1,180)) + 
  scale_color_viridis_c(option = "plasma", limits = c(1,180)) +
  geom_sf(data = pixel_spp_count_sf %>% filter(future_spp_count == 0), aes(geometry = geometry), fill = "gray40", color = "grey40") + 
  labs(title = "VoCC future number of bird species",
       subtitle = "(0 species in gray)",
       fill = "count",
       color = "count")
```

```{r}
plot_grid(historic_spp_num + theme(legend.position = 'bottom'), 
          future_spp_num + theme(legend.position = 'bottom'))
```

## Take a closer look at Florida

```{r}
fl_sf <- pixel_spp_count_sf %>% 
  filter(juris == "Florida")
```

```{r}
# Get FL pixel IDs
fl_pixels <- fl_sf$pixel_id
```

```{r}
# Filter the presence dataframe for florida pixels
fl_presence <- pixel_spp_list_presence %>% 
  filter(pixel_id %in% fl_pixels)
```

```{r}
# Look at one of the species that mostly is not in historic but is in future FL
# [27] "
hist_27 <- hist_rast[[27]]
plot(hist_27, main = names(hist_rast)[27])
```

```{r}
# Test out one of the maps
future_27 <- future_rast_vocc[[27]]
plot(future_27, main = names(future_rast_vocc)[27])
```

```{r}
# Look at one of the species that mostly is not in historic but is in future FL
# [31] "
hist_31 <- hist_rast[[31]]
plot(hist_31, main = names(hist_rast)[31])
```

```{r}
# Test out one of the maps
future_31 <- future_rast_vocc[[31]]
plot(future_31, main = names(future_rast_vocc)[31])
```


### Both species above look like they have one cell in FL in historic and then that range expands up through the channel (so lots of species get concentrated in that channel)

## Let's map only the species that have at least one cell in FL in historic

```{r}
fl_spp_hist <- fl_presence %>% 
  filter(historical == 1)

fl_spp_hist_vec <- unique(fl_spp_hist$spp)
```

```{r}
fl_spp_hist_all <- pixel_spp_list_presence %>% 
  filter(spp %in% fl_spp_hist_vec)
```

```{r}
# Count up species by pixel
fl_pixel_spp_count <- fl_spp_hist_all %>% 
  group_by(pixel_id, juris) %>% 
  summarize(historic_spp_count = sum(historical, na.rm = TRUE),
            future_spp_count = sum(future, na.rm = TRUE))
```

```{r}
# Merge in species count
fl_pixel_spp_count_sf <- inner_join(id_rast_sf, fl_pixel_spp_count)
```

```{r}
fl_historic_spp_num <- ggplot() +
  geom_sf(data = fl_pixel_spp_count_sf, aes(geometry = geometry, fill = historic_spp_count, color = historic_spp_count)) + 
  scale_fill_viridis_c(option = "plasma", limits = c(1,180)) + 
  scale_color_viridis_c(option = "plasma", limits = c(1,180)) +
  #geom_sf(data = fl_pixel_spp_count_sf %>% filter(historic_spp_count == 0), aes(geometry = geometry), fill = "gray40", color = "grey40") + 
  labs(title = "Historic number of bird species (FL start)",
       subtitle = "(0 species in gray)",
       fill = "count",
       color = "count")
```

```{r}
fl_future_spp_num <- ggplot() +
  geom_sf(data = fl_pixel_spp_count_sf, aes(geometry = geometry, fill = future_spp_count, color = future_spp_count)) + 
  scale_fill_viridis_c(option = "plasma", limits = c(1,180)) + 
  scale_color_viridis_c(option = "plasma", limits = c(1,180)) +
  #geom_sf(data = fl_pixel_spp_count_sf %>% filter(future_spp_count == 0), aes(geometry = geometry), fill = "gray40", color = "grey40") + 
  labs(title = "Future number of bird species (FL start)",
       subtitle = "(0 species in gray)",
       fill = "count",
       color = "count")
```

```{r}
plot_grid(fl_historic_spp_num + theme(legend.position = 'bottom'), 
          fl_future_spp_num + theme(legend.position = 'bottom'))
```

## How do these same species compare to the Lawler data? 

```{r}
# Read in Lawler data
lawler_spp_mapfiles <- list.files(here_anx('/data_lawler_2020/species_projections'), 
                           full.names = TRUE, recursive = TRUE, pattern = '.tif')
lawler_spp_all_df <- data.frame(f = lawler_spp_mapfiles,
                         spp = basename(lawler_spp_mapfiles) %>%
                           str_remove('(_gf85|_in85|_mc85|_bias).+') %>%
                           str_replace_all('_', ' '),
                         tax = basename(dirname(lawler_spp_mapfiles)))
```

```{r}
# Filter for just Lawler birds
birds_all_df <- lawler_spp_all_df %>% 
  filter(tax == "Birds")
```

```{r}
birds_all <- birds_all_df %>% 
  group_by(spp, tax) %>% 
  summarize(files = n())
```

For each species in the sample, aggregate the three future scenarios into a single map.  Crop to the lower continental US states.

```{r}
### Aggregate over the vector of species names
lawler_spp_vec <- birds_all_df$spp %>% unique()

lawler_future_rast <- lapply(lawler_spp_vec, FUN = agg_scenario, df = birds_all_df) %>%
  setNames(lawler_spp_vec) %>%
  rast()

#x <- future_rast[[1]]
#plot(x, main = names(future_rast)[1])
```

### crop historical maps

```{r}
lawler_hist_rast <- lapply(lawler_spp_vec, FUN = agg_scenario, df = birds_all_df, 
                    scenario = 'historical') %>%
  setNames(lawler_spp_vec) %>%
  rast()

x_lawler <- lawler_future_rast[[1]]
plot(x_lawler, main = names(lawler_future_rast)[1])

```

## Generate jurisdiction map

Let's generate a simple jurisdiction map - US states, and Canada and Mexico.

```{r}
states_sf <- ne_states(returnclass = 'sf', country = c('Mexico', 'Canada', 'United States of America')) 

state_id_df <- states_sf %>%
  st_drop_geometry() %>%
  mutate(juris = ifelse(adm0_a3 == 'USA', name, admin)) %>%
  dplyr::select(id = gn_id, juris)
### plot(states_sf %>% select(iso_a2))
lawler_states_rast <- rasterize(states_sf, lawler_hist_rast, field = 'gn_id')
#plot(lawler_states_rast)
```

## Rough pass at classification?

For each spp let's create a dataframe of jurisdictions and cell values...

```{r}
lawler_spp_list <- lapply(lawler_spp_vec, FUN = extract_jurisdictions_pixels,
                         future_r = lawler_future_rast, 
                         hist_r = lawler_hist_rast,
                         juris_r = lawler_states_rast) %>%
  setNames(lawler_spp_vec) %>%
  bind_rows(.id = 'spp') %>%
  left_join(state_id_df, by = 'id') #%>%
  ### note, Canada and Mexico have multiple jurisdictions (provinces and states)
  ### so sum up all those instances
  #group_by(spp, juris) %>%
  #summarize(n_historical = sum(n_historical),
            #n_future = sum(n_future), 
            #.groups = 'drop')
```

Filter lawler_spp_list for the same set of FL birds

```{r}
lawler_spp_list_fl <- lawler_spp_list %>% 
  mutate(spp = tolower(spp)) %>% 
  filter(spp %in% fl_spp_hist_vec) %>% 
    filter(!is.na(id)) %>%
    filter(!is.na(historical) | !is.na(future)) %>% 
  # Assign future pixels > 0.5 to spp prescence 
  mutate(future = case_when(future > 0.5 ~ 1,
                           TRUE ~ 0))
```

```{r}
lawler_fl_count <- lawler_spp_list_fl %>% 
  group_by(pixel_id, juris) %>% 
  summarize(historic_spp_count = sum(historical, na.rm = TRUE),
            future_spp_count = sum(future, na.rm = TRUE))
```

```{r}
# Create a blank raster with same extent as x = hist_rast[[1]] 
id_rast_lawler <- rast(x_lawler)
```

```{r}
# Assign IDs and then convert to sf
values(id_rast_lawler ) <- 1:ncell(x_lawler)
```

```{r}
id_rast_sf_lawler <- id_rast_lawler %>% 
  stars::st_as_stars() %>% 
  st_as_sf() %>% 
  rename("pixel_id" = "Acrocephalus familiaris")
```

```{r}
# Merge in count
lawler_fl_count_sf <- inner_join(id_rast_sf_lawler, lawler_fl_count)
```

```{r}
lawler_fl_historic_spp_num <- ggplot() +
  geom_sf(data = lawler_fl_count_sf, aes(geometry = geometry, fill = historic_spp_count, color = historic_spp_count)) + 
  scale_fill_viridis_c(option = "plasma", limits = c(1,180)) + 
  scale_color_viridis_c(option = "plasma", limits = c(1,180)) +
  #geom_sf(data = fl_pixel_spp_count_sf %>% filter(historic_spp_count == 0), aes(geometry = geometry), fill = "gray40", color = "grey40") + 
  labs(title = "Lawler historic number of bird species (FL start)",
       subtitle = "(0 species in gray)",
       fill = "count",
       color = "count")
```

```{r}
lawler_fl_future_spp_num <- ggplot() +
  geom_sf(data = lawler_fl_count_sf, aes(geometry = geometry, fill = future_spp_count, color = future_spp_count)) + 
  scale_fill_viridis_c(option = "plasma", limits = c(1,180)) + 
  scale_color_viridis_c(option = "plasma", limits = c(1,180)) +
  #geom_sf(data = fl_pixel_spp_count_sf %>% filter(future_spp_count == 0), aes(geometry = geometry), fill = "gray40", color = "grey40") + 
  labs(title = "Lawler future number of bird species (FL start)",
       subtitle = "(0 species in gray)",
       fill = "count",
       color = "count")
```

```{r}
plot_grid(lawler_fl_historic_spp_num + theme(legend.position = 'bottom'), 
          lawler_fl_future_spp_num + theme(legend.position = 'bottom'))
```

```{r}
# Look at one of the species that mostly is not in historic but is in future FL
# [27] "
lawler_hist_27 <- lawler_hist_rast[[27]]
plot(lawler_hist_27, main = names(lawler_hist_rast)[27])
```

```{r}
# Test out one of the maps
lawler_future_27 <- lawler_future_rast[[27]]
plot(lawler_future_27, main = names(lawler_future_rast)[27])
```

```{r}
# Look at one of the species that mostly is not in historic but is in future FL
# [31] "
lawler_hist_31 <- lawler_hist_rast[[31]]
plot(lawler_hist_31, main = names(lawler_hist_rast)[31])
```

```{r}
# Test out one of the maps
lawler_future_31 <- lawler_future_rast[[31]]
plot(lawler_future_31, main = names(lawler_future_rast)[31])
```

## What if we focus on the Lawler species that leave FL? How do these compare to the VoCC maps?

```{r}
lawler_fl_species <- lawler_spp_list %>% 
  filter(juris == "Florida") %>%
  mutate(spp = tolower(spp)) %>% 
  group_by(spp) %>% 
  summarize(historical_pixel_count = sum(historical > 0.5),
            future_pixel_count = sum(future > 0.5)) %>% 
  filter(historical_pixel_count > 0 & future_pixel_count == 0) %>% 
  ungroup()

lawler_left_fl <- lawler_fl_species$spp
```

```{r}
lawler_spp_list_left_fl <- lawler_spp_list %>% 
  mutate(spp = tolower(spp)) %>% 
  filter(spp %in% lawler_left_fl ) %>% 
    filter(!is.na(id)) %>%
    filter(!is.na(historical) | !is.na(future)) %>% 
  # Assign future pixels > 0.5 to spp prescence 
  mutate(future = case_when(future > 0.5 ~ 1,
                           TRUE ~ 0))
```

```{r}
lawler_left_fl_count <- lawler_spp_list_left_fl %>% 
  group_by(pixel_id, juris) %>% 
  summarize(historic_spp_count = sum(historical, na.rm = TRUE),
            future_spp_count = sum(future, na.rm = TRUE))
```

```{r}
# Merge in count
lawler_left_fl_count_sf <- inner_join(id_rast_sf_lawler, lawler_left_fl_count)
```

```{r}
lawler_left_fl_historic_spp_num <- ggplot() +
  geom_sf(data = lawler_left_fl_count_sf, aes(geometry = geometry, fill = historic_spp_count, color = historic_spp_count)) + 
  scale_fill_viridis_c(limits = c(1,53)) + 
  scale_color_viridis_c(limits = c(1,53)) +
  #geom_sf(data = fl_pixel_spp_count_sf %>% filter(historic_spp_count == 0), aes(geometry = geometry), fill = "gray40", color = "grey40") + 
  labs(title = "Lawler historic number of bird species (left FL)",
       subtitle = "(0 species in gray)",
       fill = "count",
       color = "count")
```

```{r}
lawler_left_fl_future_spp_num <- ggplot() +
  geom_sf(data = lawler_left_fl_count_sf, aes(geometry = geometry, fill = future_spp_count, color = future_spp_count)) + 
  scale_fill_viridis_c(limits = c(1,53)) + 
  scale_color_viridis_c(limits = c(1,53)) +
  #geom_sf(data = fl_pixel_spp_count_sf %>% filter(future_spp_count == 0), aes(geometry = geometry), fill = "gray40", color = "grey40") + 
  labs(title = "Lawler future number of bird species (left FL)",
       subtitle = "(0 species in gray)",
       fill = "count",
       color = "count")
```

```{r}
plot_grid(lawler_left_fl_historic_spp_num + theme(legend.position = 'bottom'), 
          lawler_left_fl_future_spp_num + theme(legend.position = 'bottom'))
```

## Filter the VoCC data for the same 77 species that left FL in the Lawler data

```{r}
left_fl_spp_hist_all <- pixel_spp_list_presence %>% 
  filter(spp %in% lawler_left_fl)
```

```{r}
# Count up species by pixel
left_fl_pixel_spp_count <- left_fl_spp_hist_all %>% 
  group_by(pixel_id, juris) %>% 
  summarize(historic_spp_count = sum(historical, na.rm = TRUE),
            future_spp_count = sum(future, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
# Merge in species count
left_fl_pixel_spp_count_sf <- inner_join(id_rast_sf, left_fl_pixel_spp_count)
```

```{r}
left_fl_historic_spp_num <- ggplot() +
  geom_sf(data = left_fl_pixel_spp_count_sf, aes(geometry = geometry, fill = historic_spp_count, color = historic_spp_count)) + 
  scale_fill_viridis_c(limits = c(1,53)) + 
  scale_color_viridis_c(limits = c(1,53)) +
  #geom_sf(data = fl_pixel_spp_count_sf %>% filter(historic_spp_count == 0), aes(geometry = geometry), fill = "gray40", color = "grey40") + 
  labs(title = "VoCC historic number of bird species (left FL)",
       subtitle = "(0 species in gray)",
       fill = "count",
       color = "count")
```

```{r}
left_fl_future_spp_num <- ggplot() +
  geom_sf(data = left_fl_pixel_spp_count_sf, aes(geometry = geometry, fill = future_spp_count, color = future_spp_count)) + 
  scale_fill_viridis_c(limits = c(1,53)) + 
  scale_color_viridis_c(limits = c(1,53)) +
  #geom_sf(data = fl_pixel_spp_count_sf %>% filter(future_spp_count == 0), aes(geometry = geometry), fill = "gray40", color = "grey40") + 
  labs(title = "VoCC future number of bird species (left FL)",
       subtitle = "(0 species in gray)",
       fill = "count",
       color = "count")
```

```{r}
plot_grid(left_fl_historic_spp_num + theme(legend.position = 'bottom'), 
          left_fl_future_spp_num + theme(legend.position = 'bottom'))
```

```{r}
plot_grid(lawler_left_fl_historic_spp_num + theme(legend.position = 'right'), 
          lawler_left_fl_future_spp_num + theme(legend.position = 'right'),
          left_fl_historic_spp_num + theme(legend.position = 'right'), 
          left_fl_future_spp_num + theme(legend.position = 'right'))
```

### For comparison, create the same four grid map forallthe birds

```{r}
lawler_spp_list_all <- lawler_spp_list %>% 
  mutate(spp = tolower(spp)) %>% 
    filter(!is.na(id)) %>%
    filter(!is.na(historical) | !is.na(future)) %>% 
  # Assign future pixels > 0.5 to spp prescence 
  mutate(future = case_when(future > 0.5 ~ 1,
                           TRUE ~ 0))
```

```{r}
lawler_count <- lawler_spp_list_all%>% 
  group_by(pixel_id, juris) %>% 
  summarize(historic_spp_count = sum(historical, na.rm = TRUE),
            future_spp_count = sum(future, na.rm = TRUE))
```

```{r}
# Merge in count
lawler_count_sf <- inner_join(id_rast_sf_lawler, lawler_count)
```

```{r}
lawler_historic_spp_num <- ggplot() +
  geom_sf(data = lawler_count_sf, aes(geometry = geometry, fill = historic_spp_count, color = historic_spp_count)) + 
  scale_fill_viridis_c(option = "plasma", limits = c(1,180)) + 
  scale_color_viridis_c(option = "plasma", limits = c(1,180)) +
  #geom_sf(data = fl_pixel_spp_count_sf %>% filter(historic_spp_count == 0), aes(geometry = geometry), fill = "gray40", color = "grey40") + 
  labs(title = "Lawler historic number of bird species",
       subtitle = "(0 species in gray)",
       fill = "count",
       color = "count")
```

```{r}
lawler_future_spp_num <- ggplot() +
  geom_sf(data = lawler_count_sf, aes(geometry = geometry, fill = future_spp_count, color = future_spp_count)) + 
  scale_fill_viridis_c(option = "plasma", limits = c(1,180)) + 
  scale_color_viridis_c(option = "plasma", limits = c(1,180)) +
  #geom_sf(data = fl_pixel_spp_count_sf %>% filter(future_spp_count == 0), aes(geometry = geometry), fill = "gray40", color = "grey40") + 
  labs(title = "Lawler future number of bird species",
       subtitle = "(0 species in gray)",
       fill = "count",
       color = "count")
```

```{r}
plot_grid(lawler_historic_spp_num + theme(legend.position = 'right'), 
          lawler_future_spp_num + theme(legend.position = 'right'),
          historic_spp_num + theme(legend.position = 'right'), 
          future_spp_num + theme(legend.position = 'right'))
```

