---
title: "IUCN + spp crossing borders - bird traits"
format: 
  html:
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(here)
library(iucnredlist) ### remotes::install_github("IUCN-UK/iucnredlist")
library(taxize)
```

## Gather maps and aggregate

Following the method in 2_combine_spp_scenarios.qmd: 
Grab a sample of species (10 spp for each taxon), gather all future scenario maps and aggregate, summing and dividing by 3 to make a "probability" map for future distributions, and crop to just the lower 48 US states (and a little Canada and Mexico).  Then crop the historical maps to the same extent.  

### Identify species sample

```{r}
here_anx <- function(f = '', ...) { 
  ### create file path to git-annex dir for project
  f <- paste(f, ..., sep = '/')
  f <- stringr::str_replace_all(f, '\\/+', '/')
  f_anx <- sprintf('/home/shares/usgs-cap-rangeshifts/data_lawler_2020/%s', f)
  return(f_anx)
}
```

List all map files available, and gather info on species, taxon, and scenario from the file path information.

```{r}
spp_mapfiles <- list.files(here_anx('species_projections'), 
                           full.names = TRUE, recursive = TRUE, pattern = '.tif')
spp_all_df <- data.frame(f = spp_mapfiles,
                         spp = basename(spp_mapfiles) %>%
                           str_remove('(_gf85|_in85|_mc85|_bias).+') %>%
                           str_replace_all('_', ' '),
                         tax = basename(dirname(spp_mapfiles)))

### sample 100 birds (301 total), and identify all scenario maps for each
set.seed(20)
spp_sample_df <- spp_all_df %>%
  filter(tax == "Birds") %>% 
  select(-f) %>%
  distinct() %>%
  group_by(tax) %>%
  slice_sample(n = 100) %>%
  ungroup() %>%
  left_join(spp_all_df)
```

### Aggregate future projections

For each species in the sample, aggregate the three future scenarios into a single map.  Crop to the lower continental US states.

```{r}
### set up extent for continental US
us_extent <- ext(c(xmin = -130, xmax = -65, ymin = 23, ymax = 50))

### Define aggregation function
agg_scenario <- function(s, df, scenario = 'future') {
  ### for species s, filter historical map files from df; rasterize; mean
  
  if(scenario == 'future') { ### filter to non-Historical
    s_df <- df %>%
      filter(spp == s & !str_detect(f, 'Historical'))
  } else { ### else filter to Historical
    s_df <- df %>%
      filter(spp == s & str_detect(f, 'Historical'))
  }
  s_map <- rast(s_df$f) %>%
    crop(us_extent) %>%
    sum(na.rm = TRUE)
  s_map <- s_map / nrow(s_df)
  return(s_map)
}

### Aggregate over the vector of species names
spp_vec <- spp_sample_df$spp %>% unique()

future_rast <- lapply(spp_vec, FUN = agg_scenario, df = spp_sample_df) %>%
  setNames(spp_vec) %>%
  rast()

x <- future_rast[[1]]
#plot(x, main = names(future_rast)[1])
```

### crop historical maps

```{r}
hist_rast <- lapply(spp_vec, FUN = agg_scenario, df = spp_sample_df, 
                    scenario = 'historical') %>%
  setNames(spp_vec) %>%
  rast()

x <- hist_rast[[1]]
#plot(x, main = names(hist_rast)[1])

```

## Generate jurisdiction map

Let's generate a simple jurisdiction map - US states, and Canada and Mexico.

```{r}
states_sf <- ne_states(returnclass = 'sf', country = c('Mexico', 'Canada', 'United States of America')) 

state_id_df <- states_sf %>%
  st_drop_geometry() %>%
  mutate(juris = ifelse(adm0_a3 == 'USA', name, admin)) %>%
  select(id = gn_id, juris)
### plot(states_sf %>% select(iso_a2))
states_rast <- rasterize(states_sf, hist_rast, field = 'gn_id')
#plot(states_rast)
```

## Rough pass at classification?

For each spp let's create a dataframe of jurisdictions and cell values...

```{r}
extract_jurisdictions <- function(spp, future_r, hist_r, juris_r) {
  ### spp <- spp_vec[1]
  ### future_r = future_rast; hist_r = hist_rast; juris_r = states_rast
  s_f_r <- future_r[[names(future_r) == spp]]
  s_h_r <- hist_r[[names(hist_r) == spp]]
  s_df <- data.frame(values(s_h_r), 
                     values(s_f_r), 
                     values(juris_r)) %>%
    setNames(c('historical', 'future', 'id')) %>%
    filter(!is.na(id)) %>%
    filter(!is.na(historical) | !is.na(future)) %>%
    group_by(id) %>%
    summarize(n_historical = sum(historical > .25),
              n_future = sum(future > .5))
  
  return(s_df)
}

state_spp_list <- lapply(spp_vec, FUN = extract_jurisdictions,
                         future_r = future_rast, 
                         hist_r = hist_rast,
                         juris_r = states_rast) %>%
  setNames(spp_vec) %>%
  bind_rows(.id = 'spp') %>%
  left_join(state_id_df, by = 'id') %>%
  ### note, Canada and Mexico have multiple jurisdictions (provinces and states)
  ### so sum up all those instances
  group_by(spp, juris) %>%
  summarize(n_historical = sum(n_historical),
            n_future = sum(n_future), 
            .groups = 'drop')
```

## Interpret results

Interpret historical and future cell counts into entry/exit/no presence/continued presence.  NOTE: probably would want to throw a threshold on the cell counts, like at least 10 cells to actually register...

```{r}
all_juris_df <- state_spp_list %>%
  group_by(spp) %>% 
  mutate(range_historical = sum(n_historical),
         range_future = sum(n_future),
         range_ratio = range_future/range_historical,
         prec_historical = 100*n_historical/sum(n_historical),
         prec_future = 100*n_future/sum(n_future),
         prec_pt_change = prec_future - prec_historical) %>% 
  mutate(status = case_when(n_historical == 0 &  n_future > 0 ~ 'entry into new',
                            n_historical > 0  & n_future == 0 ~ 'exit from historical',
                            n_historical == 0 & n_future == 0 ~ 'no presence',
                            TRUE ~ 'continued presence')) %>% 
  ungroup()

all_juris_df$status %>% table()
```

```{r}
entry_exit <- all_juris_df %>% 
  filter(status %in% c("entry into new", "exit from historical")) #%>% 
  #separate(spp, into = c("genus", "species"), sep = " ", remove = FALSE) %>% 
  #rename(scientific_name = spp)
```

```{r}
# Just focusing on country level jurisdiction movement for now
entry_exit_ca_mx <- entry_exit %>% 
  mutate(country = case_when(juris == "Canada" ~ "Canada",
                             juris == "Mexico" ~ "Mexico",
                             TRUE ~ "United States")) %>% 
  filter(country != "United States")

spp_vec_ca_mx <- unique(entry_exit_ca_mx$spp)
```

```{r}
# Run species names through taxize
temp <- gna_verifier(spp_vec_ca_mx) %>% 
  separate(currentCanonicalSimple, into = c("current_genus", "current_species"), sep = " ", remove = FALSE)
```

```{r}
# Pull the genus and species names
gns_vec_current <- temp$current_genus
species_vec_current <- temp$current_species
```

## Add IUCN information

For those species that enter into a new area or exit an area, add IUCN information.
Based on code in 3_explore_spp_threat_status.qmd

```{r}
api_key <- Sys.getenv("IUCN_API_KEY") ### usethis::edit_r_environ() to add key to R environment file
api <- init_api(api_key)
```

```{r}
# Write a function to get the most recent, global iucn assessment id for each species
iucn_id <- function(api, genus, species){
  # get species info 
 species_info <- assessments_by_name(api, genus, species) 
 list_status <- species_info %>% 
   filter(scopes_code == 1 & latest == TRUE) #%>% 
   #filter(year_published == max(year_published))
 return(list_status$assessment_id)
}
```

```{r}
# Muscicapa dauurica was orignial species name and correct name for IUCN
species_vec_current[21] <- "dauurica"

# Taxize didn't catch mispelling of Pica nutalli
species_vec_current[29] <- "nutalli"
```


```{r}
# Loop through function, save as list
iucn_id_list <- list()
for(i in 1:length(species_vec_current)){
  test <- iucn_id(api, gns_vec_current[i], species_vec_current[i])
  iucn_id_list[[i]] <- test
}
```

```{r}
id_vector <- unlist(iucn_id_list)
```

```{r}
# Run the assessment numbers collected from the iucn_id function
id_data <- assessment_data_many(api, id_vector, wait_time = 0.5)
```

```{r}
# Extract the IUCN status from all the sample species (or any other variable of interest)
# Run element_names() to see what you can extract
id_names <- extract_element(id_data, "taxon")
# taxon_synonyms might be able to help with some of the scientific name issues
id_synonyms <- extract_element(id_data, "taxon_synonyms")
id_iucn_status <- extract_element(id_data, "red_list_category")
id_pop_trends <- extract_element(id_data, "population_trend")

id_names_status <- left_join(id_names, id_iucn_status, by = "assessment_id")
id_names_status_trends <- left_join(id_names_status, id_pop_trends, by = "assessment_id")
```

```{r}
# Merge in entry / exit information 
# Get the crosswalk of lawler names and taxize names
spp_taxize_xwlk <- temp %>% 
  select(submittedName, currentCanonicalSimple) %>% 
  # Correct the current_genus, current_species names for which taxize didn't work
  mutate(scientific_name = case_when(currentCanonicalSimple == "Muscicapa latirostris" ~ submittedName,
                                     currentCanonicalSimple == "Pica nuttalli" ~ "Pica nutalli",
                                     TRUE ~ currentCanonicalSimple)) %>% 
  separate(scientific_name, into = c("genus", "species"), sep = " ", remove = FALSE)

id_names_status_trends_submittednme <- left_join(id_names_status_trends, spp_taxize_xwlk, by = "scientific_name")

entry_exit_ca_mx_clean <- entry_exit_ca_mx %>% 
  rename(submittedName = spp)

entry_exit_iucn <- inner_join(entry_exit_ca_mx_clean , id_names_status_trends_submittednme, by = "submittedName")
```

# What about seeing which species are only native to the three countries?

```{r}
id_loc <- extract_element(id_data, "locations")
```

```{r}
# 10 species
us_endemic <- id_loc %>% 
  # Only country to which species are endemic is the US for this batch of species 
  filter(is_endemic == TRUE)

# endemic assessment ids
endemic_ids <- unique(us_endemic$assessment_id)
```

# Filter the birds for the endemic species
```{r}
entry_exit_iucn_endemic <- entry_exit_iucn %>% 
  filter(assessment_id %in% endemic_ids)
```
```{r}
plot(hist_rast$`Ammodramus maritimus`, main = "Ammodramus maritimus - historical")
plot(future_rast$`Ammodramus maritimus`, main = "Ammodramus maritimus - future")
```

## Try merging in the AVONET traits data 
# https://figshare.com/s/b990722d72a26b5bfead 

```{r}
library(janitor)
# Merge in AVONET dataset (reading in BirdLife tab, which should match with IUCN names)
avonet <- read_csv(here("data-laa", "AVONET-birdlife-tab-supplementary-dataset-1.csv")) %>% 
  clean_names()
```

```{r}
entry_exit_endemic_clean <- entry_exit_iucn_endemic %>% 
  select(scientific_name, juris:sis_id, description.x, description.y) %>% 
  rename(redlist_status = description.x, 
         pop_trend = description.y)
```

```{r}
entry_exit_endemic_traits <- left_join(entry_exit_endemic_clean, (avonet %>% rename(scientific_name = species1)))
```

# Next steps - take a look at the trait data that was merged in 

How well do the names in the Avon trait data match with the Lawler bird names? Avon trait data should have the IUCN names that will run through their api. 

```{r}
birds_all_df <- spp_all_df %>% 
  filter(tax == "Birds")
```

```{r}
birds_all <- birds_all_df %>% 
  group_by(spp, tax) %>% 
  summarize(files = n())
```

```{r}
# See how well the Avonet trait data merges in 
# 270 of 301 have matches - could work with these species for now since they should be able to match wiht IUCN
birds_avonet <- left_join(birds_all, (avonet %>% rename(spp = species1)))
```

```{r}
lawler_avvonet_birds <- birds_avonet %>% 
  drop_na(sequence)

lawler_avvonet_birds_vec <- lawler_avvonet_birds$spp
```

```{r}
# Select the 270 birds from the dataset that have avonet matches 
spp_birds_df <- spp_all_df %>%
  filter(spp %in% lawler_avvonet_birds_vec) %>% 
  left_join(spp_all_df)
```

```{r}
### Aggregate over the vector of species names
birds_future_rast <- lapply(lawler_avvonet_birds_vec, FUN = agg_scenario, df = spp_birds_df) %>%
  setNames(lawler_avvonet_birds_vec) %>%
  rast()

#x <- future_rast[[1]]
#plot(x, main = names(future_rast)[1])

```

### crop historical maps

```{r}
birds_hist_rast <- lapply(lawler_avvonet_birds_vec, FUN = agg_scenario, df = spp_birds_df, 
                    scenario = 'historical') %>%
  setNames(lawler_avvonet_birds_vec) %>%
  rast()

#x <- birds_hist_rast[[1]]
#plot(x, main = names(hist_rast)[1])

```

```{r}
bird_state_spp_list <- lapply(lawler_avvonet_birds_vec, FUN = extract_jurisdictions,
                         future_r = birds_future_rast, 
                         hist_r = birds_hist_rast,
                         juris_r = states_rast) %>%
  setNames(lawler_avvonet_birds_vec) %>%
  bind_rows(.id = 'spp') %>%
  left_join(state_id_df, by = 'id') %>%
  ### note, Canada and Mexico have multiple jurisdictions (provinces and states)
  ### so sum up all those instances
  group_by(spp, juris) %>%
  summarize(n_historical = sum(n_historical),
            n_future = sum(n_future), 
            .groups = 'drop')
```

## Interpret results

Interpret historical and future cell counts into entry/exit/no presence/continued presence.  NOTE: probably would want to throw a threshold on the cell counts, like at least 10 cells to actually register...

```{r}
birds_all_juris_df <- bird_state_spp_list %>%
  group_by(spp) %>% 
  mutate(range_historical = sum(n_historical),
         range_future = sum(n_future),
         range_ratio = range_future/range_historical,
         prec_historical = 100*n_historical/sum(n_historical),
         prec_future = 100*n_future/sum(n_future),
         prec_pt_change = prec_future - prec_historical) %>% 
  mutate(status = case_when(n_historical == 0 &  n_future > 0 ~ 'entry into new',
                            n_historical > 0  & n_future == 0 ~ 'exit from historical',
                            n_historical == 0 & n_future == 0 ~ 'no presence',
                            TRUE ~ 'continued presence')) %>% 
  ungroup()

birds_all_juris_df$status %>% table()
```

note - could do some of the trait data stats with the table above (and then also add in IUCN data)

```{r}
birds_all_juris_avonet <- left_join(birds_all_juris_df, (avonet %>% rename(spp = species1)))
```
```{r}
# entry/exit stats (can see if statistically different later)
# also averaging all the jurisdictions together (maybe should only include each bird once)
entry_exit_bird_stats <- birds_all_juris_avonet %>% 
  group_by(status) %>% 
  summarise(mean_mass = mean(mass),
            mean_tarsus_length = mean(tarsus_length))

entry_exit_bird_stats 
```


