---
title: "Species change at pixel level"
format: 
  html:
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(here)
library(iucnredlist) ### remotes::install_github("IUCN-UK/iucnredlist")
library(janitor)
```

# See how species change at the pixel level

```{r}
# Functions 
source(here("common_fxns.R"))
```

```{r}
# Read in Lawler data
spp_mapfiles <- list.files(here_anx('/data_lawler_2020/species_projections'), 
                           full.names = TRUE, recursive = TRUE, pattern = '.tif')
spp_all_df <- data.frame(f = spp_mapfiles,
                         spp = basename(spp_mapfiles) %>%
                           str_remove('(_gf85|_in85|_mc85|_bias).+') %>%
                           str_replace_all('_', ' '),
                         tax = basename(dirname(spp_mapfiles)))
```

```{r}
# Read in AVONET dataset (reading in BirdLife tab, which should match with IUCN names)
avonet <- read_csv(here("data-laa", "AVONET-birdlife-tab-supplementary-dataset-1.csv")) %>% 
  clean_names()
```

```{r}
# Filter for just Lawler birds
birds_all_df <- spp_all_df %>% 
  filter(tax == "Birds")
```

```{r}
birds_all <- birds_all_df %>% 
  group_by(spp, tax) %>% 
  summarize(files = n())
```

## Gather maps and aggregate

Following the method in 2_combine_spp_scenarios.qmd: 
For all the Lawler birds species (301) gather all future scenario maps and aggregate, summing and dividing by 3 to make a "probability" map for future distributions, and crop to just the lower 48 US states (and a little Canada and Mexico).  Then crop the historical maps to the same extent. 

### Aggregate future projections

For each species in the sample, aggregate the three future scenarios into a single map.  Crop to the lower continental US states.

```{r}
### set up extent for continental US
us_extent <- ext(c(xmin = -130, xmax = -65, ymin = 23, ymax = 50))

### Aggregate over the vector of species names
spp_vec <- birds_all_df$spp %>% unique()

future_rast <- lapply(spp_vec, FUN = agg_scenario, df = birds_all_df) %>%
  setNames(spp_vec) %>%
  rast()

x <- future_rast[[1]]
#plot(x, main = names(future_rast)[1])
```

### crop historical maps

```{r}
hist_rast <- lapply(spp_vec, FUN = agg_scenario, df = birds_all_df, 
                    scenario = 'historical') %>%
  setNames(spp_vec) %>%
  rast()

x <- hist_rast[[1]]
#plot(x, main = names(hist_rast)[1])

```

## Generate jurisdiction map

Let's generate a simple jurisdiction map - US states, and Canada and Mexico.

```{r}
states_sf <- ne_states(returnclass = 'sf', country = c('Mexico', 'Canada', 'United States of America')) 

state_id_df <- states_sf %>%
  st_drop_geometry() %>%
  mutate(juris = ifelse(adm0_a3 == 'USA', name, admin)) %>%
  dplyr::select(id = gn_id, juris)
### plot(states_sf %>% select(iso_a2))
states_rast <- rasterize(states_sf, hist_rast, field = 'gn_id')
#plot(states_rast)
```

```{r}
# Function to get pixels instead of jurisdictions
extract_jurisdictions_pixels <- function(spp, future_r, hist_r, juris_r) {
  ### spp <- spp_vec[1]
  ### future_r = future_rast; hist_r = hist_rast; juris_r = states_rast
  s_f_r <- future_r[[names(future_r) == spp]]
  s_h_r <- hist_r[[names(hist_r) == spp]]
  s_df <- data.frame(values(s_h_r),
                     values(s_f_r),
                     values(juris_r)) %>%
    setNames(c('historical', 'future', 'id')) %>%
    mutate(pixel_id = row_number()) %>% 
    filter(!is.na(id)) %>%
    filter(!is.na(historical) | !is.na(future)) #%>%
    #group_by(id) %>%
    #summarize(n_historical = sum(historical > .25),
              #n_future = sum(future > .5))

  return(s_df)
}
```

## Rough pass at classification?

For each spp let's create a dataframe of jurisdictions and cell values...

```{r}
state_spp_list <- lapply(spp_vec, FUN = extract_jurisdictions_pixels,
                         future_r = future_rast, 
                         hist_r = hist_rast,
                         juris_r = states_rast) %>%
  setNames(spp_vec) %>%
  bind_rows(.id = 'spp') %>%
  left_join(state_id_df, by = 'id') #%>%
  ### note, Canada and Mexico have multiple jurisdictions (provinces and states)
  ### so sum up all those instances
  #group_by(spp, juris) %>%
  #summarize(n_historical = sum(n_historical),
            #n_future = sum(n_future), 
            #.groups = 'drop')
```


# Now that we have each species at the pixel level (pixel id), how many species are in each pixel for historical and future?

```{r}
pixel_spp <- state_spp_list %>% 
  group_by(pixel_id, juris) %>% 
  summarize(n_spp_historical = sum(historical > .25),
            n_spp_future = sum(future > .5)) %>% 
  mutate(n_spp_diff = n_spp_future - n_spp_historical)

```

```{r}
# Vast majority of pixels are losing total # of species (simply from numbers perspective)
ggplot(data = pixel_spp) +
  geom_histogram(aes(x = n_spp_diff),
                 bins = 50) +
  labs(title = "Bird species gain from historic to future by pixel",
       x = "species change")
```

# Next step - sorensen similarity index (test a few pixels)
